---
title: "Reproducible Research: Peer Assessment 1"
author: Achilleas Katsaros
output: 
  html_document:
    keep_md: true
---

```{r setoptions, include=FALSE}
# include libraries and set default options, not visible in the HTML document
library(dplyr)
library(knitr)
opts_chunk$set(fig.path = "figure/")
```

## Loading and preprocessing the data

For this purpose, it is enough to unzip the compressed input file and read the CSV.

```{r}
activity_data <- read.csv(unzip('activity.zip'))
```

## What is mean total number of steps taken per day?

First, a `steps_per_day` data frame is created, with the help of the `dplyr` package.

```{r}
# group the data frame by date and sum the number of steps, removing NA's
by_date <- group_by(activity_data, date)
steps_per_day <- summarize(by_date, sum_steps = sum(steps, na.rm=TRUE))
```

This is the histogram of the total number of steps for each day.

```{r}
hist(steps_per_day$sum_steps, breaks = 10, main = 'Histogram of total steps per day', xlab = 'Total steps per day', col = 'green')
```

```{r}
# calculate the mean and median total number of steps per day
mean_val <- round(mean(steps_per_day$sum_steps), 2)
median_val <- median(steps_per_day$sum_steps)
```

The **mean** and **median** total number of steps per day are `r mean_val` and `r median_val` respectively.

## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
